---
title: "traffic_fatalities_sf_15_17"
author: "Jeremie Young"
date: "November 6, 2018"
output: html_document
---

# load accident dataset as CSV file
```{r}
library(readr)
accident2015 -> read_csv("C:/Users/jeremiey/Google Drive/Springboard Intro to Data Science/Capstone/Raw Datasets/accident2015.csv")
accident2016 <- read_csv("C:/Users/jeremiey/Google Drive/Springboard Intro to Data Science/Capstone/Raw Datasets/accident2016.csv")
accident2017 <- read_csv("C:/Users/jeremiey/Google Drive/Springboard Intro to Data Science/Capstone/Raw Datasets/accident2017.csv")
```

# import both dplyr and tidyr packages to use
```{r}
library(dplyr)
library(tidyr)
```

# merge datasets
```{r}
accident15_17 <- bind_rows(accident2015, accident2016, accident2017)
```



# subset observations of the dataset to CA state (6) AND San Francisco County codes (1333 - FORT MASON MILITARY RESERVATION (HISTORICAL), 1337 - FORT WINFIELD SCOTT, 2905 - PRESIDIO OF SAN FRANCISCO (HISTORICAL), 3290 - SAN FRANCISCO
```{r}
accident15_17 <- accident15_17 %>% 
             filter(CITY == 3290 | CITY == 1333 | CITY == 1337 | CITY == 2905)
```

# Are 116 local observations enough to create an analysis? If not, I would collect more datasets from previous years
# there are additional variables that need to be removed from the newer dataframes (2015-2017) to match with the older ones
# some outputs have been slightly changed within a those variables (they are number coded)
# 



